---
title: "Class08 - Halloween"
author: "Assael Madrigal (PID: A10179083)"
format: pdf
---

load the file
```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candies

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
#we can also look at which ones are fruity with logicals
candy[ as.logical(candy$fruity),]

```
There are 38 types of fruity candy

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Snickers",]$winpercent
```
Snickers has a win percent of 76%

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
Kit kat also has a winpercent of 76%

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
Tootsie rolls has a 49% winpercent

```{r}
#library("skimr")
#this way i don't have to load the whole package, just the function
#skimr::skim
#skim_candy <- skim(candy)
#skim_candy
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

All of the data has a value between 0 and 1 but the winrate is between 0 and 100, so it would dominate the PCA.
So we need to **scale** it

> Q7. What do you think a zero and one represent for the candy$chocolate column?

0 means that the candy has no chocolate, 1 it means it has it

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)

library(ggplot2)
ggplot(candy, aes(winpercent)) +
  geom_histogram(bins=10)
```

> Q9. Is the distribution of winpercent values symmetrical?

Not perfectly symetrical, but very close to it so it does not deviate too much and the center looks very close to the median.

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```

The center of distribution is slightly below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy[as.logical(candy$chocolate),"winpercent"])
mean(candy[as.logical(candy$fruity),"winpercent"])
mean(candy[as.logical(candy$chocolate),"winpercent"]) > mean(candy[as.logical(candy$fruity),"winpercent"])
```
Yes it is, chocolate is 61% and fruity is 44%

>Q12. Is this difference statistically significant?

```{r}
t.test(candy[as.logical(candy$chocolate),"winpercent"], candy[as.logical(candy$fruity),"winpercent"])
```
Yes, it is significant because the p-value is less than 0.05

> Q13 What are the five least liked candy types in this set?

```{r}
candy[order(candy$winpercent),]
head(candy[order(candy$winpercent),], n=5)

```
The worst 5 are Nik, Boston, chiclets, super and jawbusters

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```
The top 5 favorites are reeses, reeses, twix, kit kat and snickers

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

Color a vector by if it has chocolates
```{r}
#rep() just makes a repeat of what we say plus the number of times
mycols <- rep("gray", nrow(candy))

mycols[ as.logical(candy$chocolate)] <- "blue"


#applying this to candy
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets are ranked the worst chocolate
> Q18. What is the best ranked fruity candy?

Starburst is ranked the best fruity candy

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
library(dplyr)

candy_mutated <- candy %>%
  mutate(bang_per_buck = pricepercent/winpercent)%>%
  arrange(bang_per_buck)

```
Tootsie Roll is the best candy for bang per buck 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Top most expensive are NikLNip, Ring pop, Nestle Smarties, Mily way and Simply caramel
of these the least popular is NikLNip

> Optional Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}

ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)

corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and chocolate are the most anti-correlated
> Q23. Similarly, what two variables are most positively correlated?

The most correlated is winpercent and chocolate with a corelation of 0.63

```{r}
pca <- prcomp(candy, scale. = TRUE)
summary(pca)
```

```{r}
#plot(pca$x[,1:2])
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The most contribution for PC1 is fruity, then pluribus then hard
